Think deep for reef resilience

### 2025-09-25 paper/paper_draft.qmd
Add Pedro's content to supplementary and reference to recent Botelho et al calibrated model paper

An exploratory scenario modelling approach was adopted for this investigation. The method
involves exploring a wide range of potential futures under different assumptions of
environmental (climatic) and ecological conditions. Exploratory modelling approaches have
gained favour in recent years to explore the implications of deep uncertainty; where there
is little or no agreement on the key drivers of a system, the extent of the relationship,
how best to represent them or their probability distributions [@conradt2024; @lempert2013].
The approach embraces uncertainty as an inherent characteristic of complex environmental
systems [@lempert2010; @lempert2013] and is a core methodology underpinning the development
of robust and adaptive management strategies [@haasnoot2013].

@@

\![Number of years reef coral cover exceeds the assumed carbonate budget threshold in terms
of coral cover ($\theta_0$) between 2025 to 2075 (y-axis) under ACCESS-ESM1-5. Left panel
shows relationship between reef median depth, while the right panel shows the same for
log10-weighted incoming connectivity values. Point colour (purple to yellow) represents
($\theta_0$), with brighter shades indicating a higher $\theta_0$. Deeper reefs have
increased number of years above the considered range of $\theta_0$, while no relationship
between weighted incoming connectivity and carbonate budget maintenance was found.
(../figs/ACCESS-ESM1-5/depth_conn_carbonate_comparison.png){#fig-access-esm1-5-depth-conn-carbonate
fig-align="center" width="14.82cm"}

@@

The modelling conducted with ADRIA-CoralBlox itself is subject to many uncertainties. For
one, a formal mechanistic and spatial validation of the model represents an area for future
refinement and is the focus of a future publication. Indicative results are included in the
supplementary **\[TI: For Pedro to include! :`D`\].** The consistency of projected patterns
with known coral-climate relationships and reef characteristics provides confidence in the
model's ability to meaningfully capture ecological signals at the GBR-scale. The mechanistic
basis of the model, combined with qualitative assessment of model behavior, supports the
robustness of our findings for understanding broad-scale coral cover responses to climatic
drivers. The aim here, however, was to assess the relative influence of reef characteristics
on reef resilience under future conditions, rather than provide spatially explicit and
precise predictions of reef outcomes. Using this approach, we suggest that reef depth is an
important factor for reef resilience under thermal stress.

@@

>>>>

An exploratory scenario modelling approach was adopted for this investigation. The method
involves exploring a wide range of potential futures and model parameterisations under
different assumptions of environmental (climatic) and ecological conditions. Exploratory
modelling approaches have gained favour in recent years to explore the implications of deep
uncertainty; where there is little or no agreement on the key drivers of a system, the
extent of the relationship, how best to represent them or their probability distributions
[@conradt2024; @lempert2013]. The approach embraces uncertainty as an inherent
characteristic of complex environmental systems [@lempert2010; @lempert2013] in an attempt
to address limitations of calibrated ecosystem models with a singular set of parameter
values [@lubianabotelho2025]. Exploratory scenario modelling is a core methodology
underpinning the development of robust and adaptive management strategies [@haasnoot2013].

@@

![Number of years reef coral cover exceeds the assumed carbonate budget threshold in terms
of coral cover ($\theta_0$) between 2025 to 2075 (y-axis) under ACCESS-ESM1-5. Left panel
shows relationship between reef median depth, while the right panel shows the same for
log10-weighted incoming connectivity values. Point colour (purple to yellow) represents
($\theta_0$), with brighter shades indicating a higher $\theta_0$. Deeper reefs have
increased number of years above the considered range of $\theta_0$, while no relationship
between weighted incoming connectivity and carbonate budget maintenance was
found.](../figs/ACCESS-ESM1-5/depth_conn_carbonate_comparison.png){#fig-access-esm1-5-depth-conn-carbonate
fig-align="center" width="14.82cm"}

@@

The modelling conducted with ADRIA-CoralBlox itself is subject to many uncertainties. For
one, a formal mechanistic and spatial validation of the model represents an area for future
refinement and is the focus of a future publication. The consistency of projected patterns
with known coral-climate relationships and reef characteristics provides confidence in the
model's ability to meaningfully capture ecological signals at the GBR-scale. The mechanistic
basis of the model supports the robustness of our findings for understanding broad-scale
coral cover responses to climatic drivers. The aim here, however, was to assess the relative
influence of reef characteristics on reef resilience across a range of future conditions and
parameterisations, rather than provide spatially explicit and precise predictions of reef
outcomes. Using this approach, we suggest that reef depth is an important factor for reef
resilience under thermal stress.

@@

### Calibration

The growth and mortality rate parameters of ADRIA-CoralBlox were calibrated using hard coral
cover and benthic composition data from the Long-Term Monitoring Program (LTMP). From all
the reefs for which there is LTMP data available, 99 were selected for having at least 4
observations that span more than 10 years. These locations were randomly split into a
calibration set (with 73 reefs) and a validation set (with 26 reefs), as evenly spatially
distributed as possible. The error function used in the calibration process is composed of
three elements [@eq-calibration]. The first is a Mean Absolute Exponential Error ($MAEE$)
that captures the difference between modeled and observed hard coral cover. The second is a
Coral Composition Error ($CCE$) that captures the differences in the benthic composition
between modeled and observed coral cover. The third ($Err~p~$) is an error that emphasizes
differences between observed and modelled cover at the beginning and end of the series, as
well as on peaks and troughs:

$$
err = MAEE_\sigma_  *  w_1_  +  CEE_t_  *  w_c_  +  Err_p_  *  w_p_
$$ {#eq-calibration}

[@sfig-reef-scale-calibration] shows a comparison between modeled and observed relative
coral cover for 6 reefs with distinct disturbance patterns. For each one, a Root Mean Square
Error (RMSE), the MAEE and Spearman’s Rank Correlation Coefficient (SRCC) is also shown.
[@sfig-man-area-calibration] presents another comparison between observed and modeled hard
coral cover aggregated by GBRMPA management area.

![Time series comparing modelled and observed relative coral cover for selected validation
reefs. These reefs were selected to show distinct disturbance patterns and behaviors. reefs
(a) and (b), with low heat stress and a single disturbance at the beginning, show the
model’s ability of reproducing the reef’s recovery under low or no disturbance; reefs (c)
and (d), with moderate to high heat stress and disturbance events in the middle of the
considered timeframe, show the model’s ability to reproduce disturbances reasonably well but
some difficulty in representing the post-disturbance recovery ; reefs (e) and (f) illustrate
the challenge of reproducing the observed behavior in a high disturbance
scenario.](../figs/reef_scale_calibration.png){#sfig-reef-scale-calibration
fig-align="center" width="14.82cm"}

![Comparison between observed and modelled coral relative cover aggregated, for each GBRMPA
management area, for the calibration set (first row) and validation set (second row). For
each one the aggregated RMSE and MAE are
presented.](../figs/management_area_calibration.png){#sfig-man-area-calibration
fig-align="center" width="14.82cm"}
